---
main_topsize: 0.15
main_bottomsize: 0.025
title: '**A case study using the *rerddap* R package**'
author:
  - name: '**Virginia A. García Alonso**'
    affil: 1,2
    main: true
    orcid: '0000-0003-2224-8376'
    twitter: VA_GarciaAlonso 
    email: garciaalonso.v.a@gmail.com
  - name: Pablo S. Milla Carmona
    affil: 3,4
    main: true
    orcid: '0000-0002-0396-6244'
    twitter: millacarmonia
    email: 'pablomillac@gmail.com'
affiliation:
  - num: 1
    address: Universidad de Buenos Aires. Facultad de Ciencias Exactas y Naturales. Departamento de Biodiversidad y Biología Experimental. Buenos Aires, Argentina.
  - num: 2
    address: CONICET-Universidad de Buenos Aires. Instituto de Biodiversidad y Biología Experimental Aplicada (IBBEA). Buenos Aires, Argentina.
  - num: 3
    address: Universidad de Buenos Aires. Facultad de Ciencias Exactas y Naturales. Departamento de Ciencias Geológicas. Buenos Aires, Argentina.
  - num: 4
    address: CONICET-Universidad de Buenos Aires. Instituto de Estudios Andinos "Don Pablo Groeber" (IDEAN). Buenos Aires, Argentina.
main_findings:
  - "Retrieving and visualizing **satellite sea water temperature** data for marine analyses"
output: 
  posterdown::posterdown_betterport:
    self_contained: false
    pandoc_args: --mathjax
    number_sections: false
bibliography: packages.bib
link-citations: true
affiliation_textsize: "25pt"
reference_textsize: "25pt"
body_textsize: "36pt"
#knit: pagedown::chrome_print
---
```{css, echo = FALSE}
@page {
size: $if(poster_width)$$poster_width$$else$36in$endif$ $if(poster_height)$$poster_height$$else$48in$endif$;
margin: 0;
padding: 0;
}
body {
margin: 0px;
padding: 0px;
width: $if(poster_width)$$poster_width$$else$36in$endif$;
height: $if(poster_height)$$poster_height$$else$48in$endif$;
text-align: justify;
font-size: $if(body_textsize)$$body_textsize$$else$45px$endif$;
line-height: 1.05;
}
/* RMarkdown Class Styles */
/* center align leaflet map,
from https://stackoverflow.com/questions/52112119/center-leaflet-in-a-rmarkdown-document */
.html-widget {
margin: auto;
position: sticky;
margin-top: 2cm;
margin-bottom: 2cm;
}
.leaflet.html-widget.html-widget-static-bound.leaflet-container.leaflet-touch.leaflet-fade-anim.leaflet-grab.leaflet-touch-drag.leaflet-touch-zoom {
position: sticky;
width: 100%;
}
pre.sourceCode.r, pre.sourceCode.python {
background-color: transparent;
border: solid $if(secondary_colour)$$secondary_colour$$else$#008080$endif$ 1mm;
border-radius: 0.5em;
padding: 0mm;
margin: 0;
margin-top: 0.5em;
margin-bottom: 0.5em;
/* align-items: center; */
}
div.sourceCode {
background-color: #00000010;
width: 100%;
margin-left: 0.5em;
margin-right: 0.5em;
}
code.sourceCode.r, code.sourceCode.python {
background-color: transparent;
font-size: 27pt;
border-radius: 2mm;
}
code {
font-size: 27pt;
font-family: monospace;
background-color: $if(secondary_colour)$$secondary_colour$24$else$#00808024$endif$;
color: $if(secondary_colour)$$secondary_colour$$else$#008080$endif$;
padding: 0.5mm;
line-height: 1;
border-radius: 1mm;
}
caption {
margin-bottom: 10px;
font-size: 35pt;
font-style: italic;
}

tbody tr:nth-child(odd) {
    background-color: $if(secondary_colour)$$secondary_colour$$else$#008080$endif$20;
}
.table>thead>tr>th, .table>tbody>tr>th, .table>tfoot>tr>th, .table>thead>tr>td, .table>tbody>tr>td, .table>tfoot>tr>td{
  border-spacing: 0;
  font-size: 80%;
  border-style: none;
  padding-top: 15px;
  padding-bottom: 15px;
  padding-right: 1em;
  padding-left: 1em;
  line-height: 1em;
}
table {
  margin: auto;
}
th {
  padding-left: 5mm;
  padding-right: 5mm;
}
.caption {
font-size: 30pt;
font-style: italic;
padding-top: 0;
}
.references {
font-size: $if(reference_textsize)$$reference_textsize$$else$20px$endif$;
line-height: 90%;
}
/* Create three unequal columns that floats next to each other */
.column {
float: left;
padding: 0px;
}
.outer {
width: $if(poster_width)$$poster_width$$else$36in$endif$;
height: calc($if(poster_height)$$poster_height$$else$48in$endif$ * $if(main_topsize)$ (1 - $main_topsize$ - $main_bottomsize$ - 0.01) $else$ 0.6325 $endif$);
-webkit-column-count: $if(column_numbers)$$column_numbers$$else$3$endif$; /* Chrome, Safari, Opera */
-moz-column-count: $if(column_numbers)$$column_numbers$$else$3$endif$; /* Firefox */
column-count: $if(column_numbers)$$column_numbers$$else$3$endif$;
-webkit-column-fill: auto;
-moz-column-fill: auto;
column-fill: auto;
column-gap: 0;
padding-left: 0cm;
padding-right: 0cm;
/* -webkit-column-rule-width: 50%;
-moz-column-rule-width: 50%;
column-rule-width: 50%; */
-webkit-column-rule-style: none;
-moz-column-rule-style: none;
column-rule-style: none;
-webkit-column-rule-color: black;
-moz-column-rule-color: black;
column-rule-color: black;
background-color: $if(body_bgcol)$$body_bgcol$$else$#ffffff$endif$;
font-family: $if(font_family)$$font_family$$else$Rasa$endif$;
margin-top: calc($if(poster_height)$$poster_height$$else$48in$endif$ * $if(main_topsize)$ $main_topsize$ $else$ 0.25 $endif$);
padding-top: 1em;
padding-bottom: 1em;
}
span.citation {
  color: $if(secondary_colour)$$secondary_colour$$else$#008080$endif$;
  font-weight: bold;
}
a {
text-decoration: none;
color: $if(secondary_colour)$$secondary_colour$$else$#008080$endif$;
}
#title {
font-size: $if(title_textsize)$$title_textsize$$else$125pt$endif$;
text-align: left;
margin: 0;
line-height: 98%;
border-bottom: 0;
font-weight: normal;
background: 0;
}
#author {
color:$if(author_textcol)$$author_textcol$$else$$if(primary_colour)$$primary_colour$$else$#0b4545$endif$$endif$;
margin: 0;
line-height: 85%;
font-size: $if(author_textsize)$$author_textsize$$else$1.17em$endif$;
}
#affiliation {
padding-top: 0.1em;
color: $affiliation_textcol$;
font-style: italic;
font-size: $if(affiliation_textsize)$$affiliation_textsize$$else$25px$endif$;
margin: 0;
}
sup {
color: $if(accent_colour)$$accent_colour$$else$#cc0000$endif$;
}
.affiliation sup {
font-size: 20px;
}
.author {
text-align: left;
}
.author sup {
font-size: 30px;
}
.author_extra {
color: $if(secondary_colour)$$secondary_colour$$else$#008080$endif$;
margin: 0;
line-height: 85%;
font-size: $if(authorextra_textsize)$$authorextra_textsize$$else$35px$endif$;
text-align: left;
}
.outer h1, h2, h3, h4, h5, h6 {
text-align: center;
margin: 0;
font-weight: bold;
}
.section h1 {
  text-align:center;
  padding-bottom:5px;
  background:
    linear-gradient(
      to left,
      $if(body_bgcol)$$body_bgcol$$else$#ffffff$endif$ 1%,
      $if(body_bgcol)$$body_bgcol$$else$#ffffff$endif$ 20%,
      $if(primary_colour)$$primary_colour$$else$#0b4545$endif$75 33%,
      $if(primary_colour)$$primary_colour$$else$#0b4545$endif$ 50%,
      $if(primary_colour)$$primary_colour$$else$#0b4545$endif$75 66%,
      $if(body_bgcol)$$body_bgcol$$else$#ffffff$endif$ 80%,
      $if(body_bgcol)$$body_bgcol$$else$#ffffff$endif$ 99%
    )
    left
    bottom
    $if(body_bgcol)$$body_bgcol$$else$#ffffff$endif$
    no-repeat;
  background-size:100% 5px ;
  margin-top: 0.5em;
  margin-bottom: 0.5em;
}
.outer h2 {
text-align: center;
}
.outer p, .level2 {
color: $if(body_textcol)$$body_textcol$$else$#000000$endif$;
}
.outer ol {
padding-left: 8%;
padding-right: 8%;
text-align: left;
}
.main {
width: $if(poster_width)$$poster_width$$else$36in$endif$;
height: calc($if(poster_height)$$poster_height$$else$48in$endif$ * $if(main_topsize)$$main_topsize$$else$0.25$endif$);
position: absolute;
background-color: $if(primary_colour)$$primary_colour$$else$#0b4545$endif$;
color: $if(main_textcol)$$main_textcol$$else$#ffffff$endif$$if(main_textalpha)$$main_textalpha$$else$90$endif$;
font-family: $if(main_fontfamily)$$main_fontfamily$$else$Special Elite$endif$;
background-image: linear-gradient($if(primary_colour)$$primary_colour$$else$#0b4545$endif$ 50%,$if(secondary_colour)$$secondary_colour$$else$#008080$endif$);
}
.main strong {
color: $if(main_textcol)$$main_textcol$$else$#ffffff$endif$;
}
.main strong > sup {
color: $if(main_textcol)$$main_textcol$$else$#ffffff$endif$;
}
.main sup {
color: $if(main_textcol)$$main_textcol$$else$#ffffff$endif$$if(main_textalpha)$$main_textalpha$$else$90$endif$;
}
#main-img-left {
width: 10%;
left: 0.5in;
bottom: 0.2in;
position: absolute;
}
#main-img-center {
width: 10%;
left: calc($if(poster_width)$$poster_width$$else$36in$endif$ * 0.45);
bottom: 0.5in;
position: absolute;
}
#main-img-right {
width: 10%;
right: 0.5in;
bottom: 0.2in;
position: absolute;
}
.main p {
font-size: $if(main_textsize)$$main_textsize$$else$150px$endif$;
font-family: $if(main_fontfamily)$$main_fontfamily$$else$Special Elite$endif$;
text-align: $if(main_textalign)$$main_textalign$$else$center$endif$;
margin: 0;
position: absolute;
top: 50%;
-ms-transform: translateY(-50%);
transform: translateY(-50%);
margin-left: 1em;
}
.fab {
color: $if(affiliation_textcol)$$affiliation_textcol$$else$#00000060$endif$;
font-size: 25px;
}
.twitter, i {
color: #00000090;
font-size: 35px;
text-decoration: none;
}
a.email {
text-decoration: none;
color: #00000090;
font-size: 35px;
}
.envelope {
color: $if(affiliation_textcol)$$affiliation_textcol$$else$#00000060$endif$;
font-size: 5px;
text-decoration: none;
}
.globe {
color: $if(affiliation_textcol)$$affiliation_textcol$$else$#00000060$endif$;
font-size: 5px;
text-decoration: none;
}
.globetext {
color: $if(affiliation_textcol)$$affiliation_textcol$$else$#00000060$endif$;
font-size: 35px;
text-decoration: none;
}
.poster_wrap {
width: $if(poster_width)$$poster_width$$else$36in$endif$;
height: $if(poster_height)$$poster_height$$else$48in$endif$;
padding: 0cm;
}
.main_bottom {
width: $if(poster_width)$$poster_width$$else$36in$endif$;
height: calc($if(poster_height)$$poster_height$$else$48in$endif$ * $if(main_bottomsize)$$main_bottomsize$$else$0.1$endif$);
margin-top: calc($if(poster_height)$$poster_height$$else$48in$endif$ * $if(main_bottomsize)$(1 - $main_bottomsize$)$else$0.9$endif$);
position: absolute;
background-color: $if(primary_colour)$$primary_colour$$else$#0b4545$endif$;
background-image: linear-gradient($if(secondary_colour)$$secondary_colour$$else$#008080$endif$ 10%, $if(primary_colour)$$primary_colour$$else$#0b4545$endif$);
}
.section {
  padding-left: $if(column_padding)$$column_padding$$else$10mm$endif$;
  padding-right: $if(column_padding)$$column_padding$$else$10mm$endif$;
}
span > #tab:mytable {
  font-weight: bold;
}
.orcid img {
  width: 3%;
}
.emphasis {
  background-color: #008080;
  color: #ffffff;
  border: solid #0b2045 3mm;
  margin: 1em;
  padding-left: 0;
  padding-right: 0;
}
.emphasis h1 {
  font-weight: bold;
  background: none;
  background-color: #0b2045;
  padding-bottom: 5mm;
  padding-top: 1mm;
  margin-top: -1mm;
  margin-right: -1mm;
  margin-left: -1mm;
}
.emphasis blockquote {
  border: 0;
}
.emphasis ol {
  padding: 0;
  padding-left: 8%;
  font-size: 100%;
  font-weight: bold;
}
.emphasis p {
  color: #ffffff;
}
```
```{r setup, include=FALSE}
#to suppress the version number in the subdirectory name
options(htmltools.dir.version = FALSE)

#to change the default values of chunk options 
knitr::opts_chunk$set(echo = FALSE, #prevent printing of the R code
                      warning = FALSE, #prevent printing of warnings
                      message = FALSE, #prevent printing of messages
                      tidy = FALSE,
                      fig.align = 'center',
                      out.width = "100%",
                      dev = 'svg', #set figures' formats
                      dev.args = list(bg = "transparent"))
options(knitr.table.format = "html") 

#to generate references for R packages
knitr::write_bib(c('rerddap','tidyverse','sf','ggspatial','rnaturalearth','rnaturalearthdata','rnaturalearthhires','marmap'), 'packages.bib')
```

# Motivation

Environmental variables are key determinants of many biological processes in marine ecosystems. Temperature variability is especially important in high latitudes where organisms are subject to marked seasonal variations which influence their life cycles. Since collection of *in situ* data entails different logistic challenges and may provide inadequate spatiotemporal resolutions, **satellite data emerges as a powerful tool to boost marine analyses**.


## What is this poster about?

A **workflow** including the necessary steps to i) retrieve **satellite sea surface temperature (sst) data** from the ERDDAP server using the *rerddap* package [@R-rerddap] and ii) **visualize** the results.


## Where and when?

The chosen study area is located in the southern border of the Southwest Atlantic Ocean, a region displaying both a **marked seasonality** and a **longitudinal gradient** in water temperature across seasons. We will visualize sst data from austral spring and autumn of 2015-2016.


## Which data set?

The 'jplMURSST41' gridded data set, a **Multiscale Ultrahigh Resolution (MUR) L4 analysis** of sst. It includes a **global 0.01 degree grid** with interpolated sst expressed as Celsius degrees (ºC). More information can be found at the [podaac dataset website](<https://podaac.jpl.nasa.gov/dataset/MUR-JPL-L4-GLOB-v4.1>).

To visualize other ERDDAP data sets check [this page](<https://coastwatch.pfeg.noaa.gov/erddap/griddap/index.html?page=1&itemsPerPage=1000>).


# 1. Load needed packages

Besides the *rerdapp* package, we will need the *tidyverse* [@R-tidyverse], *sf* [@R-sf], *ggspatial* [@R-ggspatial], *rnaturalearth* [@R-rnaturalearth], *rnaturalearthdata* [@R-rnaturalearthdata], *rnaturalearthhires* [@R-rnaturalearthhires] and *marmap* [@R-marmap] packages to manipulate data and visualize it nicely :)

```{r packages, echo=FALSE}
#to retrieve sea surface temperature (sst) data
library(rerddap)

#for basic data manipulation and visualization
library(tidyverse)

# to work with spatial data and improve our maps
library(sf) # st_crop()
library(ggspatial) # annotation_north_arrow() / annotation_scale()

#to retrieve world map
library(rnaturalearth) # ne_countries()
library(rnaturalearthdata) # holds data used by rnaturalearth package
library(rnaturalearthhires) # high resolution world vector maps from Natural Earth

# to retrieve bathymetric data
library(marmap) # getNOAA.bathy()

#to solve problems with the latest version of sf for st_as_sf()
sf_use_s2(FALSE)
```

# 2. Download sst data

We will retrieve sst data with the `info()` function and choose the desired data according to the spatiotemporal resolution needed with the `griddap()` function.

```{r download_sst_spring, eval=FALSE, echo=T}
#retrieve the sst data set
sstInfo <- info('jplMURSST41')

#choose the spatiotemporal resolution
sst_spring <-
  
  griddap(sstInfo,
          latitude = c(-56, -52),
          longitude = c(-70, -55),
          time = c('2015-09-21',
                   '2015-12-21'),
          fields = 'analysed_sst',
          fmt ="csv")
```

Downloading this data takes a while... you can save it as a `.csv` file to use it in the future with the following command:

`write.csv(sst_spring,"sst_spring.csv",row.names=F)`

# 3. Read and organize data

We have multiple sst for each latitude/longitude pair (one per day), so we will estimate **mean sst**.
```{r read_sst_spring, echo=T}
#load the data as following
sst_spring <- 
  
  read.csv(file="sst_spring.csv",header=T)
  
#transform and organize the data
mean_sst_spring <-
  
  #transform to a data frame
  as.data.frame(sst_spring) %>%

  #select the needed variables
  dplyr::select(longitude, latitude,
                analysed_sst) %>%
  
  #estimate mean sst for each lat/long
  group_by(longitude,latitude) %>%
  summarise(mean_sst=mean(analysed_sst)) %>%
  ungroup() %>%
  
  #set a new variable to identify the season
  mutate(season="Spring")
```

We can repeat the same procedure for **autumn** creating a new object named `mean_sst_autumn` and finally join both datasets in a single object. 

```{r download_sst_autumn, eval=FALSE}
  sst_autumn <-   
    griddap(sstInfo,
            latitude = c(-56, -51), 
            longitude = c(-70, -55),
            time = c('2016-03-21',
                     '2016-06-21'),
            fields = 'analysed_sst',
            fmt = "csv")

  #save data to use it in the future
  write.csv(sst_autumn,"sst_autumn.csv", row.names = FALSE)
```

```{r read_sst_autumn}
#load the data as following
sst_autumn <- read.csv(file = "sst_autumn.csv",
                       header = TRUE)
  
  #transform and organize data
  mean_sst_autumn <-
    
    #transform to a data frame
    as.data.frame(sst_autumn) %>%
    
    #select the needed variables
    dplyr::select(longitude,latitude,analysed_sst) %>%
    
    #estimate mean temperature for each lat/long position
    group_by(longitude,latitude) %>%
    summarise(mean_sst=mean(analysed_sst)) %>%
    ungroup() %>%
    
    #set a new variable to merge data with spring
    mutate(season = "Autumn")

```
```{r join_data, echo=T}
#join data from both seasons
mean_sst <- 
  bind_rows(mean_sst_spring,
            mean_sst_autumn)
```

# Finally, the plot!

We will plot sst data with *ggplot2*, employing the `geom_raster()` function.
```{r geom_raster, echo=T, eval=F}
ggplot()+
  
  #add sst as a raster and interpolate
  geom_raster(data = mean_sst,
              aes(x = longitude,
                  y = latitude,
                  fill = mean_sst),
                  interpolate = T)+
  
  #set the fill
  scale_fill_viridis_c(alpha = 0.85)+
  
  #separate plots for each season
  facet_grid(season~.)
```

A few more tweaks allow including the continental territory and relevant bathymetric contours to the map and getting this!

```{r basemap, echo=F, message=F, warning=F}

#base-map ----

#retrieve world map
world <- ne_countries(scale = "large", returnclass = "sf")

#subset data to the study area
study_area<-st_crop(world, xmin = -70, xmax = -55,
                           ymin = -56, ymax = -52)

#isobath ----
bathy_data <- 
  
  #get the data      
  getNOAA.bathy(lon1=-71, lon2=-54,
                lat1=-58, lat2=-51,
                #keep=T, #if the datafile is very large you can save it as a csv.
                resolution=1) %>%  #resolution in minutes
  
  #prepare data
  fortify.bathy() %>%  #extract bathymetric values
  mutate(z = as.numeric(z)) %>% 
  
  #subset for study area
  filter(z<=-1 & between(x, -70, -55) & between(y, -56, -52))


```


```{r, fig.alt = "Plot showing mean sea surface temperature data in spring 2015 and autumn 2016 in the southern region of the Southwest Atlantic Ocean. Temperature is indicated in Celsius degrees and values range between 2.60 and 6.16 degrees in spring and 3.70 and 11.03 degrees in autumn. In both seasons, temperatures are higher near the coasts of Tierra del Fuego (Argentina) and decreases towards the more oceanic area at the east of the continental shelf.", fig.height=8, out.width="100%", fig.cap = "Mean sea surface temperature (sst) in spring 2015 and autumn 2016 in the Southwest Atlantic Ocean"}

ggplot()+
  
  #add sst as a raster and interpolate
  geom_raster(data = mean_sst, 
              aes(x = longitude, 
                  y = latitude, 
                  fill = mean_sst),
                  interpolate = T)+
  
  #set the fill
  scale_fill_viridis_c(alpha = 0.85)+

  #adjust how we "see" the map
  geom_sf(data=study_area, fill= "grey20", colour = "white") +
  
  #add the 200m isobath
  geom_contour(data = bathy_data, aes(x=x, y=y, z=z),
               breaks=c(-200), size=0.5, colour="white")+
  
  #adjust the limits to the area
  coord_sf(xlim = c(-70, -55), ylim = c(-56, -52), expand = F)+
  
  #set the breaks in "x" and "y" axis
  scale_x_continuous(breaks = seq(from=-68, to=-56, by=4)) +
  scale_y_continuous(breaks = seq(from=-55, to=-53, by=2))+
  
  #add scale bar
  annotation_scale(location = "br", width_hint = 0.2, 
                   text_cex = 1, text_col= "white") +

  #add north arrow
  annotation_north_arrow(location = "br", 
                         which_north = T,
                         style = north_arrow_fancy_orienteering,
                         pad_x = unit(0.35, "cm"),
                         pad_y = unit(0.5, "cm"),
                         height = unit(2, "cm"),
                         width = unit(2, "cm"))+     

  #separate according to season
  facet_grid(forcats::fct_rev(season)~.)+
  
  #add title
  ggtitle("Seasonal sst variability") +

  #change legend title
  labs(fill = "Mean\nsst (ºC)")+

  #set theme
  theme_bw(base_size = 20)+
  
  #remove axis names and adjust strip
  theme(axis.title = element_blank(),
        legend.position = "bottom",
        legend.key.width = unit(1.5, "cm"),
        strip.background = element_blank(),
        strip.text =  element_text(face = "bold",hjust = 0))
```

Hope you find it useful!!! Materials employed for the poster are openly shared in this [GitHub repository](<https://github.com/virginiagarciaalonso/useR_2022_sst>) including a spanish version.

# References
